The Ranked Nearest Neighbour Interchange space

### 2019-06-28 paper.tex
edit proof: bounding length of path when deleting taxon

\begin{proof}
    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
    In particular, this holds for a shortest path $p$ from $T$ to $R$.
    Furthermore, deleting taxon $n+1$ and its parent $p_{n+1}$ from all trees on $p$ results in a path $p|_n$ from $T|_n$ to $R|_n$.
    All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    \todo{Is this proof sufficient? Otherwise: induction}
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    % \end{enumerate}
\end{proof}

>>>>

\begin{proof}
    Let $p$ be a shortest path from tree $T$ to $R$ and $p|_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
    This new path $p|_n$ is then a path from $T$ to $R$.
    Showing that at least $\delta := |rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
    By the following two observations one can easily see that this claim is true, indeed:
    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ with another node or an $\nni$ move on an edge incident to $v$.
    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
    Applying this to $parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.

    % Alternative: Induction:
    %
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    % \end{enumerate}
\end{proof}
